\documentclass{ximera}

\title{Surfaces and surface integrals of scalar functions}
\license{CC: 0}         % replace with an appropriate license, or set it in xmPreamble

\begin{document}

\begin{abstract}
    Surfaces and surface integrals of scalar functions
\end{abstract}

\begin{exercise}
    Find a parametrization of the portion of the sphere $x^2 + y^2 + z^2 = 9 $ above the plane $z = 3/2$. Specify the domain and the expression of the parametrization.
\end{exercise}

\begin{exercise}
    Find a parametrization of the portion of the cylinder  $x^2 + y^2 = 16 $ between the planes  $ x+ y +  z = 0$ and $2x + y - z = - 10$. Specify the domain and the expression of the parametrization.  
\end{exercise}


\begin{exercise}
    Find a parametrization of the portion of the cone  $ z = \sqrt{3 x^2 + 3 y^2 } $ outside the sphere $x^2 + y^2 + (z- 2)^2 = 4$ and inside the sphere $x^2 + y^2 + (z- 5)^2 = 25$.  Specify the domain and the expression of the parametrization.
\end{exercise}


\begin{exercise}
    Find a parametrization of the portion of the graph  $ z = \sin x + \sin y $  between the lines $x= 0$, $x = \pi$, $y = x$, and $y = \pi$. Specify the domain and the expression of the parametrization.  
\end{exercise}




\begin{exercise}
For each of the previous problems, take your parametrization 
\[    \varphi (u,v) = (x(u,v), y(u,v) , z(u,v))       \]
and consider the vector $\frac{\partial \varphi}{\partial u } \times \frac{\partial \varphi }{\partial v}$.  Identify which side of the surface it comes out of. Construct a second parametrization of the same surface where the corresponding vector points in the opposite direction. 
\end{exercise}



\begin{exercise}
    Using a parametrization and its Jacobian, find the area of the portion of the plane $5x + 3y + 4z = 60  $  in the first octant ($x, y, z \geq 0$). Assumme a metal lamina has this shape and its density is given by $\rho (x,y,z) = \sin x + 3$. Find its mass. 
\end{exercise}



\begin{exercise}
    Find the area of the portion of the paraboloid $x = 9 - y^2 - z^2$ with $x \geq - 16$.  Assume a metal lamina has this shape and its density is given by $\rho (x,y,z) = x - y  + 21$. Find its mass. 
\end{exercise}


\begin{exercise}
    Find  
    \[   \iint_{\Sigma } (z^2  +x  + 3 ) \, dS              \]
    where $\Sigma$ is the portion of the sphere $ x^2 + y^2 + z^2 = 16$ above the plane $z = -2$. 
\end{exercise}


\begin{exercise}
    Find  
    \[   \iint_{\Sigma } ( x^2 + 3 x  + 2 y^2 - 2y +  z^2  + 3 ) \, dS              \]
    where $\Sigma$ is the portion of the cone  $ z = \sqrt{3 x^2 + 3 y^2 } $ outside the sphere $x^2 + y^2 + (z- 2)^2 = 4$ and inside the sphere $x^2 + y^2 + (z- 5)^2 = 25$. 
\end{exercise}



\begin{exercise}
    Consider the surface $\Sigma \subset \mathbb{R}^3$ with parametrization 
    \[ \varphi : [0, 2 \pi ] \times [0, 2 \pi] \to \mathbb{R}^3 \] 
    given by 
    \[   \varphi (u,v) = (  (3 + \cos v) \cos u , (3 + \cos v) \sin u , \sin v   )              \]
    Describe the surface $\Sigma$ and find its area.  Compute
    \[    \iint _{\Sigma}  (  x + 2y + z^2 +3 ) \, dS             \]
\end{exercise}




\begin{exercise}
    Compute 
    \[    \iint_{\Sigma} (  xz + y   ) \, dS                        \]
    where $\Sigma \subset \mathbb{R}^3$ is the helicoid with parametrization 
    \[    \varphi (u,v) = ( u \cos v,  u \sin v  , v  )                         \]
    with $1 \leq u \leq 3$, $0 \leq v \leq 4 \pi $. 
\end{exercise}








\end{document}
