\documentclass{ximera}

\title{Divergence Theorem}
\license{CC: 0}         % replace with an appropriate license, or set it in xmPreamble

\begin{document}

\begin{abstract}
    Divergence Theorem
\end{abstract}


\section{Divergence Theorem}




\begin{exercise}
Let $\Sigma \subset \mathbb{R}^3$ be the sphere $x^2 + y^2 + z^2 = 9$, negatively oriented. Compute
\[   \iint_{\Sigma} \langle x^2 + 3z,  2y + xyz,  5z - x \rangle \cdot dS   \]
\end{exercise}

\begin{exercise}
Let $\Sigma \subset \mathbb{R}^3$ be the boundary of the box $[-1,2]\times[0,3]\times[-2,1]$, positively oriented. Compute
\[   \iint_{\Sigma} \langle 2x - yz,   3y + xz,  2yz + e^{x} \rangle \cdot dS    \]
\end{exercise}

\begin{exercise}
Let $\Sigma \subset \mathbb{R}^3$ be the boundary of the solid bounded by the cylinder $x^2 + y^2 = 1$ and the planes $z=0$ and $z=1$, oriented outward. Compute
\[    \iint_{\Sigma} \langle xz,    yz,    x^2 + y^2 \rangle \cdot dS     \]
\end{exercise}

\begin{exercise}
Let $\Sigma \subset \mathbb{R}^3$ be the ellipsoid $\dfrac{x^2}{4} + \dfrac{y^2}{9} +  z^2 = 1$, oriented outward. Compute
\[   \iint_{\Sigma} \langle 2x + yz, 3y + x \cos z,  e ^x + y^2 \rangle \cdot dS      \]
\end{exercise}

\begin{exercise}
Let $\Sigma \subset \mathbb{R}^3$ be the closed surface consisting of the portion of the paraboloid $z = 9 - x^2 - y^2$ with $z \geq 0$, together with the disk $x^2 + y^2 \leq 9$ in the $xy$-plane, oriented outward. Compute
\[    \iint_{\Sigma} \langle xy + 1,   yz - 3,   zx  + 5 y \rangle \cdot dS     \]
\end{exercise}

\begin{exercise}
Let $\Sigma \subset \mathbb{R}^3$ be the closed surface consisting of the portion of the cone $z = \sqrt{x^2 + y^2}$ below  the plane $z =  4 $ together with the disk $x^2 + y^2 \leq 16$ in the plane $z=4$, oriented outward. Compute
\[    \iint_{\Sigma} \langle xy + 3 y, 4 y +  z ,   5z + x \rangle \cdot dS      \]
\end{exercise}


\begin{exercise}
Let $\Sigma \subset \mathbb{R}^3$ be the boundary of the tetrahedron with vertices $(3,0,0)$, $(0, -3, 0)$, $(0,0,2)$, $(0,0,0)$, oriented outward. Compute
\[  \iint_{\Sigma} \langle 3x - 2y + e^z , 5x^2 - y  , z - \cos y \rangle \cdot dS    \]
\end{exercise}






\begin{exercise}
Let $\Sigma_1 \subset \mathbb{R}^3$ be the disk $x^2 + y^2 \leq 1$ in the $xy$-plane, oriented upward, and $\Sigma _2 \subset \mathbb{R}^3$ the portion of the sphere $x^2 + y^2 + z^2 = 1 $ with $z \geq 0$, oriented upward. Consider a vector field $F : \mathbb{R}^3 \to \mathbb{R}^3$ with positive divergence. Which one is larger? 
\[    \iint_{\Sigma _1} F \cdot dS \,\,\,\,\,\,\,\,\, \text{ or }  \,\,\,\,\,\,\,\,\, \iint_{\Sigma _2} F \cdot dS                       \]
\end{exercise}




\begin{exercise}
Let $\Sigma_1 \subset \mathbb{R}^3$ be the disk $y^2 + z^2 \leq 1$ in the $yz$-plane, oriented towards the $x$-axis, and $\Sigma _2 \subset \mathbb{R}^3$ the portion of the sphere $x^2 + y^2 + z^2 = 1 $ with $x \leq 0$, oriented towards the $x$-axis. Consider a vector field $F : \mathbb{R}^3 \to \mathbb{R}^3$ with negative divergence. Which one is larger? 
\[    \iint_{\Sigma _1} F \cdot dS \,\,\,\,\,\,\,\,\, \text{ or }  \,\,\,\,\,\,\,\,\, \iint_{\Sigma _2} F \cdot dS                       \]
\end{exercise}





\begin{exercise}
Give a proof of the Divergence Theorem in the case where the region is rectangular. That is, let $E = [a_1, b_1] \times [a_2, b_2] \times [a_3, b_3] \subset \mathbb{R}^3$, $\partial E$ its boundary oriented positively, and $F (x,y,z)$ a vector field whose domain contains $E$. Then
\[       \iint_{\partial E } F \cdot dS  = \iiint_E \text{div} (F) \, dV           \]
\end{exercise}






























\end{document}
