\documentclass{ximera}

\title{Surface integrals of scalar functions}
\license{CC: 0}         % replace with an appropriate license, or set it in xmPreamble

\begin{document}

\begin{abstract}
    Surface integrals of vector fields
\end{abstract}


\section{Surface integrals of vector fields}



\begin{exercise}
Let $\Sigma$ be the portion of the plane $2x + y + z = 6$ inside the cylinder $x^2 + y^2 = 4$, oriented upwards. Compute
\[ \iint_{\Sigma} \langle z, y,  3 - x \rangle \cdot dS     \]
\end{exercise}


\begin{exercise}
Let $\Sigma$ be the portion of the plane $z = 3x + 2y + 1$ between the planes $x= 0 $ and $x=1$ and between the planes $y = 0 $  and $y= 2$,  oriented upwards. Compute
\[   \iint_{\Sigma} \langle y + 2,  z - 3 ,   x + 1 \rangle \cdot dS    \]
\end{exercise}

\begin{exercise}
Let $\Sigma$ be the triangle with vertices $(1,0,0)$, $(0,2,0)$, $(0,0,3)$, oriented downwards. Compute
\[  \iint_{\Sigma} \langle x + 2,  2y - 5,  3z + 1 \rangle \cdot  dS    \]
\end{exercise}


\begin{exercise}
Let $\Sigma$ be the portion of the paraboloid $z = 4 - x^2 - y^2$ above the plane $z = 0$, oriented upwards. Compute
\[    \iint_{\Sigma} \langle x, y, 0 \rangle \cdot  dS  \]
\end{exercise}

\begin{exercise}
Let $\Sigma$ be the portion of the cone $z = \sqrt{x^2 + y^2}$ between the planes $z=1$ and $z=3$, oriented outward (away from the axis). Compute
\[  \iint_{\Sigma} \langle z, 3 + x,  1 - y \rangle \cdot dS  \]
\end{exercise}

\begin{exercise}
Let $\Sigma$ be the portion of the sphere $x^2 + y^2 + z^2 = 9$ with $z \geq 0$, oriented upward. Compute
\[   \iint_{\Sigma} \langle x,  y,  z \rangle \cdot dS   \]
\end{exercise}


\begin{exercise}
Let $\Sigma$ be the portion of the paraboloid $y = x^2 + z^2$ inside the cylinder $x^2 + z^2 = 1$, oriented in the  direction of the $y$-axis. Compute
\[   \iint_{\Sigma} \langle y ,  x + 1  , z \rangle \cdot dS   \]
\end{exercise}



\begin{exercise}
Let $\Sigma$ be the portion of the cone $x =  \sqrt{3 y^2 + 3 z^2}$ with $x \leq \frac{\sqrt{3}}{2}$ , oriented in the direction of the $x$-axis. Compute
\[   \iint_{\Sigma} \langle 2, 3x - 1 ,  y + z \rangle \cdot dS   \]
\end{exercise}






\end{document}
